---
title: "Project_2"
output:
  html_document:
    toc: true          
    toc_float: true     
    toc_depth: 3        
date: "2023-10-22"
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=F, message=FALSE, warning=FALSE}

if (!require('dplyr')){
  install.packages('dplyr')
}
if (!require('ggplot2')){
  install.packages('ggplot2')
}
if (!require('gridExtra')){
  install.packages('gridExtra')
}
if (!require('MASS')){
  install.packages("MASS")
}

if (!require('car')){
  install.packages("car")
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The MASS package contains data on housing values in the city of Boston
from the 1970s to the 1980s (data frame **Boston**). We are very
interested in looking at how **the median home value (variable medv,
measured in thousands of dollars) depends on various factors.**

The goal is to build a complete linear model, after standardizing the
predictors. Interactions between the predictors do not need to be
considered in the model.

Perform diagnostics of the model: a. Check the linearity of the
relationships b. Check for influential observations c. Check the
independence of observations d. Check for normality of the distribution
and homoscedasticity (constant variance)

Create a prediction plot of home prices based on the variable with the
largest coefficient by absolute value.

# 1 Data description

```{r, message=FALSE, warning=FALSE}
library(MASS)
df <- as.data.frame(Boston)
str(df)
```

This data frame contains the following columns:

-   **crim** per capita crime rate by town.

-   **zn** proportion of residential land zoned for lots over 25,000
    sq.ft.

-   **indus** proportion of non-retail business acres per town.

-   **chas** Charles River dummy variable (= 1 if tract bounds river; 0
    otherwise).

-   **nox** nitrogen oxides concentration (parts per 10 million).

-   **rm** average number of rooms per dwelling.

-   **age** proportion of owner-occupied units built prior to 1940.

-   **dis** weighted mean of distances to five Boston employment
    centres.

-   **rad** index of accessibility to radial highways.

-   **tax** full-value property-tax rate per \$10,000.

-   **ptratio** pupil-teacher ratio by town.

-   **black** proportion of blacks by town

-   **lstat** lower status of the population (percent).

-   **medv** median value of owner-occupied homes in \$1000s.

# 2 Building the Complete Linear Model

First, we will standardize the variables. At this initial stage, we will
not consider the interaction between the predictors.

```{r, message=FALSE, warning=FALSE}
df_standartized <-  as.data.frame(sapply(df[-4], scale))
df_standartized$chas <- as.factor(df$chas)
df_standartized$chas <-  relevel(df_standartized$chas, ref = '0')
```

Now let's build a model:

```{r,message=FALSE, warning=FALSE}
full_model <- lm(medv ~ ., data = df_standartized)
summary(full_model)
```

# 3 Model Diagnostics

## 3.1 Check for normality of the distribution and homoscedasticity (constant variance)

```{r,message=FALSE, warning=FALSE}
qqPlot(residuals(full_model), main = "QQ-Plot with Outliers Identified", ylab = 'Quantiles of the residuals distribution')
```

The quantile plot does not look very good; it cannot be said that the
standardized residuals are normally distributed.

## 3.2 Check the independence of observations

If we look at the structure of the dataset, we can notice the presence
of correlations between the predictors. Therefore, checking for
multicollinearity is especially relevant in this case.

One way to check the model for multicollinearity of the predictors is to
use VIF (Variance Inflation Factor).

If a predictor has a VIF value greater than 2, it should be excluded
from the model.

If there are several such predictors, a stepwise algorithm is applied:
calculate the VIF, remove the predictor with the highest VIF,
recalculate the VIF for the updated model, and repeat until all values
are below the threshold.

```{r,message=FALSE, warning=FALSE}
vif(full_model)
```

In this case, we can easily observe the presence of multicollinearity in
our model. Some predictors will need to be removed from the model in
further steps.

## 3.3 Check the linearity of relationships

The residuals distribution plot does not look very good. A fairly large
number of observations are beyond two standard deviations, and a clear
pattern is visible in the residuals. All of this indicates the presence
of nonlinearity in the relationship as well as heteroscedasticity
(non-constant variance).

```{r,message=FALSE, warning=FALSE}
gg_resid <- ggplot(data = full_model, aes(x = .fitted, y = .stdresid)) + 
  geom_point() + 
  geom_hline(yintercept = 0) +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 2, color = "blue") +
  geom_hline(yintercept = -2, color = "blue") +
  xlab('Predicted Value') + 
  ylab('Model Residuals')
gg_resid
```

## 3.4 Check for influential observations

```{r,message=FALSE, warning=FALSE}
cooksd <- cooks.distance(full_model)

# Create a data frame for plotting
mod_full_diag <- data.frame(observation = 1:length(cooksd), cooksd = cooksd)

# Plot Cook's distances using ggplot
ggplot(mod_full_diag, aes(x = observation, y = cooksd)) + 
  geom_bar(stat = "identity") + 
  geom_hline(yintercept = 2, color = "blue") +
  xlab('Number of observation') + 
  ylab('Cook\'s Distance')
```

None of the values exceed the conventional threshold of 2 units. There
are no influential observations.

# 4 Model Predictions

We will plot the predicted values of house prices based on the variable
with the largest absolute coefficient. In this case, it is the variable
lstat.

We cannot account for the variability of all predictors at once.
Therefore, we usually choose one that interests us the most (in our
case, lstat). Based on this, a test dataset is created, where the target
predictor takes values from the minimum to the maximum, while all other
predictors are represented by their mean values.

```{r,message=FALSE, warning=FALSE}
test_data <- data.frame(
  lstat = seq(min(df_standartized$lstat), max(df_standartized$lstat), length.out = 90),
  crim = rep(mean(df_standartized$crim), 90),
  zn = rep(mean(df_standartized$zn), 90),
  indus = rep(mean(df_standartized$indus), 90),
  nox = rep(mean(df_standartized$nox), 90),
  rm = rep(mean(df_standartized$rm), 90),
  age = rep(mean(df_standartized$age), 90),
  dis = rep(mean(df_standartized$dis), 90),
  rad = rep(mean(df_standartized$rad), 90),
  tax = rep(mean(df_standartized$tax), 90),
  ptratio = rep(mean(df_standartized$ptratio), 90),
  black = rep(mean(df_standartized$black), 90),
  chas = rep('0', 90)  # assuming 'chas' is a categorical variable
)

Predictions <- predict(full_model, newdata = test_data, interval = 'confidence')
MyData <- data.frame(test_data, Predictions)

Pl_predict <- ggplot(MyData, aes(x = lstat, y = fit)) +
  geom_ribbon(alpha = 0.2, aes(ymin = lwr, ymax = upr)) +
  geom_line() + 
  ggtitle("Multiple Linear Model Predictions") +
  xlab('Percentage of Lower Status Population (lstat)') + 
  ylab('Predicted House Price')
Pl_predict
```

# 5 Additional Section

Let’s return to our full model and recall the results of the
multicollinearity check.

```{r,message=FALSE, warning=FALSE}
vif(full_model)
```

We will iteratively remove one predictor with the highest VIF from the
model until all values are below the threshold (2).

```{r,message=FALSE, warning=FALSE}
mod2 <- update(full_model, .~. - tax)
#vif(mod2)

mod3 <- update(mod2, .~. - nox)
#vif(mod3)

mod4 <- update(mod3, .~. - dis)
#vif(mod4)

mod5 <- update(mod4, .~. - lstat)
#vif(mod5)

mod6 <- update(mod5, .~. - nox)
#vif(mod6)

mod7 <- update(mod6, .~. - rad)
#vif(mod7)

mod_good <- update(mod7, .~. - indus)
vif(mod_good)
```

The final model at this stage

When the categorical variable chas equals 0: 
medv = -7.06 - 0.11 * crim - 0.003 * zn + 7.06 * rm - 0.04 * age - 0.93 * ptratio + 0.015 * black

When the categorical variable chas equals 1: 
medv = -3.54 - 0.11 * crim - 0.003 * zn + 7.06 * rm - 0.04 * age - 0.93 * ptratio + 0.015 * black

## Step 1: Finding the Optimal Model

We can leave the model in this form, or we can try to retain only the
predictors that significantly affect the house prices (a large number of
predictors usually leads to overfitting).

### 5.1.1 Stepwise Selection of Predictors by Significance

In this work, I will use the backward selection (also known as backward
elimination) algorithm for predictor selection. As a selection
criterion, I will use the partial F-test.

```{r}
drop1(mod_good , test = 'F')
```

```{r,message=FALSE, warning=FALSE}
mod_good_unstd <- update(mod_good, .~. - zn)
drop1(mod_good_unstd, test = 'F')
```

Thus, we kept only the predictors with a significant influence on the
dependent variable ( **crim**, **rm**, **age**, **ptratio**, **black**,
**chas**.

```{r,message=FALSE, warning=FALSE}
summary(mod_good_unstd)
```

### 5.1.2 Model Diagnostics

We have already looked at several plots to diagnose our model, but for
multiple regression models, it is also necessary to create plots for
predictors that were excluded from the model. In our case, it is evident
that there are no unaccounted dependencies in the model overall, so we
do not need to return the predictors that were removed in previous
stages of the analysis.

```{r,message=FALSE, warning=FALSE}
res1 <- gg_resid + aes(x = zn)
res2 <- gg_resid + aes(x = indus)
res3 <- gg_resid + aes(x = nox)
res4 <- gg_resid + aes(x = dis)
res5 <- gg_resid + aes(x = rad)
res6 <- gg_resid + aes(x = tax)
res7 <- gg_resid + aes(x = lstat)
grid.arrange(res1, res2, res3, res4, res5, res6, res7, nrow = 4)
```

```{r,message=FALSE, warning=FALSE}
gg_resid <- ggplot(data = mod_good_unstd, aes(x = .fitted, y = .stdresid)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 2, color = "red") +
  geom_hline(yintercept = -2, color = "red") +
  xlab('Predicted Value') + 
  ylab('Standardized Residuals')

gg_resid
```

```{r,message=FALSE, warning=FALSE}
qqPlot(residuals(mod_good_unstd), xlab = 'Quantiles of Normal Distribution', ylab = 'Quantiles of Residuals Distribution')
```
The analysis showed that if any issues were corrected, it wasn't very effective. The model is far from perfect. In this case, some transformation of variables might have helped (though this is often undesirable as it complicates the interpretation of results), or the use of other statistical methods, which will likely be discussed in future projects. Additionally, attempts to account for interactions between variables within the final model did not lead to success; in fact, the Cook's distance plot became worse (not shown here). At a first approximation, we can try to predict which aspects need improvement to maximize the price of a house.

## 5.2 Customer recomendations

Finall model is:

When the categorical variable chas = 0:
medv = -7.06 - 0.11 * crim + 7.05 * rm - 0.04 * age - 0.92 * ptratio + 0.015 * black

When the categorical variable chas = 1:
medv = -3.54 - 0.11 * crim + 7.05 * rm - 0.04 * age - 0.92 * ptratio + 0.015 * black

We can see 6 parameters that, in one way or another, influence house
prices. Here are some insights that, in my opinion, should help maximize
the price of a house for sale:

-   Properties bordering the river are, on average, priced $3.52K
    higher.
-   A decrease in the per capita crime rate by one unit increases the
    house price by \$0.11K.
-   On average, each additional room increases the house price by $7.05K.
-   Reducing the proportion of residential houses built before 1940 by
    one unit increases the house price by \$0.04K. This parameter could
    be completely disregarded if building a brand-new neighborhood.
-   Reducing the pupil-to-teacher ratio in the city increases the house
    price by $0.92K.
-   The parameter black reflects the proportion of the Black population in the city, but it's unlikely to     be influenced, so no recommendations will be provided regarding this.

Thus, in the ideal neighborhood:

-   Properties border a river.
-   Low crime rate.
-   A large number of rooms in houses (up to 8 in this case, as the data
    does not show higher values).
-   It would be ideal if the number of students per class were minimal,
    for example, three.
-   The approximate house price under these conditions could be $50.1K.

